---
title: Practical multistate models
author: Terry Therneau
date: May 30, 2026
output: pdf_document
---

This was prepared as a class session for the advanced survival class in 
Waterloo.

# Introduction

Multistate time to event models based on hazards, i.e., multistate hazard models
(MSH) have been an extremely useful tool for the medical research teams that I
have been involved in at Mayo Clinic. See for instance Jack et al, PMID:35310829
and PMID: 41240917, Allen et al PMID: 41240917 and PMID: 28941364 and 
Le-Rademacher PMID: 36003089 as examples.
They are able to capture multiple aspects of a disease process, and thus can be
an important tool for deeper understanding.

This document is a "nuts and bolts" description that reprises some of those
analyses, with a focus on the data preparation, data checking, and analysis
code needed. Some mathematics will appear, but only to help clarify the
models.

For all the models, key first steps are
 * What data do we have, and what questions *could* be answered with that data?
 * The overall goal: what aspects of the process(es) do we wish to
 describe, and in relation to what predictors? No paper can do everything.
 * Lay out the multi-state model as a diagram. Refer to it often.
 * Create the analysis data set, with particular attention to questions of
   immortal time bias.
 * Look at the data (look, not just glance) and decide what to do with 
   special cases. There are invariably one or two.
   
Mathematically we describe out multistate model in terms of three processes.
 * $N_{ijk}(t)$ = the number of observed transitions from state $j$ to state $k$,
   up to time $t$ for subject $i$. 
 * $Y_{ij}(t)$ equals 1 if subject $i$ is currently in state $j$, at time $t$, 
   and at risk for transition to another state at that time. Subjects who have
   entered an absorbing state prior to $t$, e.g., death, are for instance not
   at risk for other transitions. 
 * $X_i(t)$ = the set of covariates for subject $i$ at time $t$. Most often, most
   of these covariates will remain static, e.g., treatment arm or sex, while 
   others are time-dependent, e.g. cholesterol might be remeasured at each visit.

There is a subtle difference between "at risk for an event" and "at risk for
an observed event", i.e., one that will be, if it occurs, captured in the data.
If subject $i$ has incomplete follow-up, "censoring", then
$Y_i(t)$ for observed events will become 0 sometime before they enter an 
absorbing state. In this document I will largely ignore this, and use $Y, N, X$
to refer to observed data. If censoring is not random, however, then our
estimators can be biased, and any investigation into the hows and whys of that
bias will need to carefully distinguish them.

## Immortal Time Bias

In our opinion, one of the greatest threats to valid medical research is the 
presence of immortal time bias (ITB). In short, for a valid time-to-event
model none of $N(t)$, $Y(t)$ nor $X(t)$ can contain values that depend on the
future. In our personal experience, MSH models provide, unfortunately, even
more opportunites for this error, ones which are sometimes difficult to spot.
The number of ways to create ITB appears to be near infinite. Here are a few
examples.

fill in later

## Timeline data
The survival package supports two forms of multistate data. The older and
better known one is referred to as counting process form. 
The first 3 lines below show the myeloid data in its raw form as found in
the survival package.
The next lines show the timeline form. The code to create mdata1 will come
later.

```{r, echo=FALSE}
library(survival)
table2 <- function(...) table(..., useNA="ifany")

# we will show this later
mdata1 <- tmerge(myeloid[,1:4], myeloid, id=id, death= event(futime, death),
                 cr = event(crtime), sct= event(txtime), relapse=event(rltime))

t1 <- with(myeloid, data.frame(id, trt, sex, flt3, time=0))
t2 <- with(subset(myeloid, !is.na(txtime)), data.frame(id, time=txtime,
                                                       sct=1))
mdata2 <- merge(t1, t2, all=TRUE)
t3 <- with(subset(myeloid, !is.na(crtime)), data.frame(id, time=crtime,
                                                       cr=1))
mdata2 <- merge(mdata2, t3, all=TRUE)
t4 <- with(subset(myeloid, !is.na(rltime)), data.frame(id, time=rltime,
                                                       relapse=1))
mdata2 <- merge(mdata2, t4, all=TRUE)
t5 <- with(myeloid, data.frame(id, time=futime, death))
mdata2 <- merge(mdata2, t5, all=TRUE)

indx <- match(mdata2$id, myeloid$id)
for (i in c("trt", "sex", "flt3")) mdata2[,i] <- myeloid[indx, i]
```
```{r, timeline, echo=TRUE}
subset(myeloid, id<4)

subset(mdata1, id<4)
```
Each row of data in the timeline form consists of a subject id,
a time interval (tstart, tstop], the event(s) if any that occur
(death, cr, sct, relapse), and covariates (trt, sex, flt3).
Covariate apply over the entire interval (tstart, tstop], and events
happen at the end of the interval.
In this particular data set there are no time-dependent covariates, and multiple
possible endpoints: complete response, stem cell transplant, relapse, and death.

Now the timeline form of these 3 subjects.
```{r, echo=TRUE}
subset(mdata2, id<4)
```
In timeline data each row has a unique (subject identifier, time) pair, along
with all the information that was observed at that time, with missing values
for things that were not observed. Timeline data sets have many NA values.
Static variables like treatment, sex and FLT3 state at randomization can be
filled in for every row, as was done here, or not. (The analysis code can
deal with either.)  One could argue for instance that FLT3 was
only observed at the start since a pathology assessment was not repeated,
so should be NA on row 2,
and certainly that none of these covariates were observed on the day of death.

For simple survival timeline style makes little sense: each subject will have
to have two rows, one at 0 with the covariates and another at the death/censoring
time with all covariates missing.
For more complex data they have two advantages

 * Counting process data sets often require the use of tmerge, which
   can be tricky. Timeline data can be built using more standard tools, which
   may be easier
 * If one is rigorous with respect to the "observed at this time" rule for each
   line of data, most immortal time bias problems will be avoided


# Example 1, NAFLD
The NAFLD data is found in 3 data sets. 
The original study contains all new NAFLD diagnosis in Olmsted County from 
1997 to 2014, along with 4 potential controls for each case. Controls were
selected matched on age, sex, and calendar year.

To protect patient confidentiality all time intervals are in days since the 
index date; none of the dates from the original data were retained. 
Subject age is their integer age at the index date, and the subject 
identifier is an arbitrary integer. As a final protection, the data includes
only a 90% random sample of the original data.

 * nalfd1 has one row for each subject and contains the time fixed variables
 * nafld2 contains time-dependent laboratory tests
 * nafld3 contains the outcomes
 
```{r, nafld1}
nafld1[1:5,]

nafld2[1:5,]

nafld3[1:5,]

table(nafld2$test)

table(nafld3$event)
```

The analysis in Allen 2018 looked at the accumulation of 3 metabolic
comorbidities, leading to the 4 state model shown below.
Age, male sex, NAFLD diagosis, and BMI are the available baseline covariates.
However, BMI is also a major risk factor for development of fatty liver. 
(We didn't include ongoing height/weight in the R data sets.)

```{r, nafld2}
nstate <- c("0MC", "1MC", "2MC", "3MC", "death")
nmat <- matrix(0, 5, 5, dimnames=list(nstate, nstate))
nmat[1,2] <- nmat[2,3] <- nmat[3,4] <- 1
nmat[1:4,5] <- 1
statefig(rbind(4,1), nmat)
```

In the nafld1 data set, the case.id variable, for a control, gives the
id of the case they were matched with.
For the incident cases, we set case.id = id.
The nafld3 data also has nafld diagnosis, so check the consistency.

```{r, nafld3}
ndata <- nafld1  # working data set
ndata$nafld <- 1*(ndata$id == ndata$case.id) # create a variable

table(ndata$nafld)
ndx  <- subset(nafld3, event=="nafld")
table(ndx$days==0)

subset(ndx, days>0)[1:5,]
subset(ndata, case.id == 13)

temp <- subset(ndx, days==0 & !(id %in% ndata$id[ndata$nafld==1]))
dim(temp)
temp[1:5,]
subset(ndata, id %in% c(805, 3158))
subset(ndata, case.id == 805)

```

What is going on?
 1. 318 subjects who where chosen as controls were found to have NAFLD later
 in life. You cannot exclude someone as a control at age 45 based on something
 that will happen to them some years later: this is a form of immortal time
 bias. When that control later became a case, 4 new controls for them were
 selected. 
 2. At this time NAFLD had a funny definition: if someone had fatty liver but
 also say a hepatitis C diagnosis, then they "didn't have" NAFLD. The study
 question was more or less the consquences if NAFLD was the *only* liver
 disease.
 3. What about hepatitis C 2 weeks later, 6 weeks later, ... The diagnostic
 process for someone with a GI complaint can sometimes stretch out for months,
 so the decision was that "real" NAFLD starts once 12 months has passed 
 without another smoking gun. This is day 0.
 4. Controls need to matched on this new date, otherwise the NAFLD subjects
 will benefit from immortal time bias.
 5. The 19 confuse me. Perhaps I made a mistake when going from dates
 to days? (It should be "days from inital recruitment" for all).
 
"If you don't know where you are going, you might end up someplace else". 
Yogi Berra

What should we do? Different choices lead to a different data set and analysis
changes.
 * Simple approach: only follow the orginal enrollees and their controls. Ignore
 when someone changes the time-dependent covariate to NAFLD.
 * Treat NAFLD as a time-dependent covariate
 * Treat NAFLD as a time-dependent state
 * Restart: when a case becomes a control, split them into two copies:
   an original control and a new case.


## Simple approach
For the simple analysis, we need a working data frame that has
a row/days pair each time that the total number of MC increases.
We can take advantage of the fact that each of the MC only is recorded
at most once per subject: formally the variable is "history of diabetes", etc.
The lines below are likely an inefficient way to do this, but
it works.

```{r, nafldssimple}
# first, subset to the relevant cases and controls
caseid <-  subset(ndata, id==case.id)$id
tdata1 <- subset(ndata, case.id %in% caseid)
dim(tdata1) # this is the set subjects, one row each

# the endpoints of interest
enddata <- subset(nafld3, (id %in% tdata1$id) & 
                        (event %in% c("diabetes","dyslipidemia",  "htn")))
enddata[1:10,]
table(enddata$days <=0)
temp <- with(enddata, table(days <=0, event[,drop=TRUE]))
temp
round(temp/nrow(tdata1), 2)

# counting process data
ndata2 <- tmerge(tdata1[c(1,2,3,6:7,10)], tdata1, id=id, 
                 death= event(futime, status),
                 options=list(tstartname="day1", tstopname="day2"))
ndata2 <- tmerge(ndata2, enddata, id=id, mcount=cumtdc(days), 
                 mevent= cumevent(days))
subset(ndata2, id <8)
summary(ndata2)

ndata2$istate <- factor(ndata2$mcount, 0:3, nstate[1:4])
temp <- ifelse(ndata2$death==1, 4, ndata2$mevent)
ndata2$event <- factor(temp, 0:4, c("censor", nstate[-1]))
check1 <-survcheck(Surv(day1, day2, event) ~1, data=ndata2, 
                   id=id, istate=istate)
check1

```

For an MSH model fit with survfit or coxph the underlying likelhood is
event based. 
Each observed event contributes a term to the likelihood, and presumes
 * over the window of observation for a subject, all the transitions for
   that subject are recorded
 * the window of observation is complete, i.e., it includes all times at
   which an event, if it occured, would have been recorded
 * non-informative censoring

One consequence of the first is that we do not need to inform the routine
of which transitions can occur, and which are (effectively) absorbing: it
tabulate them on its own.
But, we need to check that it's assessment agrees with our own. In this
case we have some jumps that were not in our original concept.

```{r, nfig2}
nmat <- matrix(0, 5, 5, dimnames=list(nstate, nstate))
nmat[1,2] <- nmat[2,3] <- nmat[3,4] <- 1
nmat[1:4,5] <- 1
nmat[1, 3:4] <- nmat[2,4] <- .7
statefig(rbind(4,1), nmat, alty=c(1,2,1,2,2,1,1,1,1,1))
```

To properly account for age we will do the primary analysis on age scale.

```{r, nafldage}
quantile(tdata1$age)
ndata2$age1 <- ndata2$age + ndata2$day1/365.25
ndata2$age2 <- ndata2$age + ndata2$day2/365.25

naj <- survfit(Surv(age1, age2, event) ~ male, ndata2, id=id, istate=istate,
               p0= c(1,0,0,0,0), start.time=50)

dim(naj)
plot(naj[,2:5], lwd=2, lty=1:2, col=rep(1:4, each=2), xmax=100,
     xlab="Age", ylab="Probability in state")
#plot(naj[1,], cumprob=2:5, col=1:4, yaxs='i', xmax=100,
#     xlab="Age", ylab="Probability in state")
#text(c(55, 68, 77, 79, 89), c(.8, .11, .37, .57, .69), nstate,
#     col=c(1,1,2,3,4))

dim(naj$cumhaz)
colSums(naj$n.transition)
```

The probability in state estimates are mostly oblivious to the presence 
of  0mc:2MC, 0MC:3MC and 1MC:3MC transtions, but the hazard
estimates for those will be based on very small numbers.
This affects the multi-state Cox model as well, since it is focused on hazards.

```{r, nacox1}
nacox1 <- coxph(Surv(age1, age2, event) ~ nafld + male,
                data=ndata2, id=id, istate=istate)
nacox1$cmap

nacox2 <- coxph(list(Surv(age1, age2, event) ~ nafld + male,
                     1:2 + 1:3 + 1:4 ~ nafld + male /common,
                     2:3 + 2:4 ~ nafld + male /common,
                     0:5 ~ male/common),
                data=ndata2, id=id, istate=istate)
nacox2

round(coef(nacox2, matrix=TRUE),3)

nacox3 <- coxph(list(Surv(day1, day2, event)~  nafld+ bmi + strata(case.id),
                       1:2 + 1:3 + 1:4 ~ nafld + bmi /common,
                       2:3 + 2:4 ~ nafld + bmi / common), 
                data=ndata2, id=id, istate=istate)
nacox3
```

Because all the covariates are all time fixed, absolute risk estimates are
fairly easy to get. For a Cox model, predictions are available for any
a priori combination of covariates, which we provide in a dummy data set.
Since it is multistate, we need to give a distribution of starting
states, and since we are on age scale a starting time.
(If you don't give these it will guess. I don't think I have every found the
result of that to be useful.)

```{r, absnafld}
dummy <- expand.grid(male=0:1, nafld=0:1)
coxsurv2 <- survfit(nacox2, newdata=dummy, p0=c(1,0,0,0,0), start.time=45)
dim(coxsurv2)
print(coxsurv2[1,], rmean=100)

xx <- paste0("male=", dummy$male, ", Nafld=", dummy$nafld)
sojourn <- summary(coxsurv2, rmean=100)$table
sojourn <- matrix(sojourn[,"rmean"], ncol=4, byrow=TRUE,
               dimnames=list(nstate, xx))
rbind(sojourn, Total= colSums(sojourn))


plot(coxsurv2[,-1], col=1:4, lty=rep(1:4, each=4))
```
* The mean time in state has to add to (100-45)= 55, across the states.
* Subscripting the curves is a useful tool
* The simple plot has too many curves, it looks like a bowl of spaghetti.
  Better plotting requires artistic choices.
* For a stratified model, the defalt plot will have a set of curves per
  stratum, which is not useful here.

```{r, absnafld2}
opar <- par(mfcol=c(2,2), mar=c(5,5,2,1))
xx <- outer(c("F", "M"), c("No NAFLD", "NAFLD"), paste)
for (i in 1:4) {
    plot(coxsurv2[i,], cumprob=2:5, col=1:4, lty=rep(1:2, each=4))
    text(55, .9, xx[i])
}
par(opar)

barplot(sojourn[,c(2,4,3,1)], horiz=TRUE, col=c(5:2, 0), las=2,
        names.arg=c("Male", "MaleN", "FemaleN", "Female"),
        xlab="Years in state", border=NA, legend=TRUE)
```


# Mayo Clinic Study of Aging

The MCSA is based on a stratified random sample of subjects from Olmsted
County, Minnesota. 
The dementia data set in the mshmm package is based on this data, and is
able to reproduce Jack et al (2022) quite well, so we will use it for analysis.
This data set is already in timeline form.

```{r, mcsa1}
library(mshmm)
mstate <- c("CU", "MCI", "dementia", "death")
mmat <- matrix(0, 4, 4, dimnames=list(mstate, mstate))
mmat[1,2] <- mmat[2,3] <- 1
mmat[1:3,4] <- 1
statefig(rbind(3,1), mmat)
```

```{r, mcsa2}
survcheck(Surv2(age, dx, repeated=TRUE) ~1, id=ptnum, dementia)$transitions

survcheck(Surv2(age, dx) ~1, id=ptnum, data = dementia)
```
We can see several things by looking at this table.
  * The survcheck routine was written for counting process data.
    One aspect of Surv2, which translates from timeline form to that form,
    is the repeated argument.  When a subject has two sequential visits in
    the same state, the survfit and coxph routines need to know whether to
    treat the second observation as a new event. Most of the time they
    should not, repeated infections in the CGD data set is a counterexample,
    however.  In this data each row is a visit, the repeated form clearly shows
    that over half of all visit pairs are CU:CU; most patients stay stable for
    many years before progression.  The (any) row of the total visits table
    shows that 53\% (2296/4327) have no recorded transtions during follow-up.
  * There are 30 instances of a CU:dementia pair.  Per the model we assume
    that these subjects passed through MCI, we simply did assess them often
    enough to observe the transient MCI state.
  * Dementia:MCI is much less common than MCI:CU.  This reflects on one
    hand the clinical reality that CU and MCI are more difficult to distinguish
    than dementia, and also fewer overall dementia cases.
  * There are no transitions from death to another state, or from
    death to censored (further follow-up after death).  This is as it
    should be.
  * Off study dementia only transitions to death. This is because
    further record reviews are not scheduled for demented subjects.
  * There are no transitions from nondem to nondem or off-dem. 
    This is, it turns out, an interesting artifact of how the online
    data collection system was designed.
    The nurse abstracters who review the history have a screen with fields
    for ``last assessment date for dementia' and 'dementia yes/no'.  
    At a second, third, etc review for a patient, should one occur,
    the contents of these are overwritten; thus the master data set
    at any given time only has the most recent value.
  * More dementias were found in record review than in study visits.
    We believe that this reflects informative censoring: subjects whose
    cognition is slipping selectively drop out, i.e., exactly those most
    likely to receive a dementia diagnosis at their next visit. 

To reprise the 2022 paper we need to create a new endpoint of non-demented (ND)
dementia, and death. 

```{r, nondem}
ddata <- dementia
table(ddata$dx)
temp <- c(1, 2,2,3,2,3,4)[as.numeric(ddata$dx)]
ddata$dx2 <- factor(temp, 1:4, c("censor", "ND", "dementia", "death"))
with(ddata, table(dx2, dx))

survcheck(Surv2(age, dx2) ~1, ddata, id=ptnum)
```

The 2022 analysis was actually for time to *first* dementia. This change to
using only first event is common enough that we have added an option to
the survival package.
survcheck(Surv2(age, dx2, repeated='first') ~1, ddata, id=ptnum)

```{r, first}
ddata$edu4 <-pmax(8, ddata$educ) /4
check2 <- survcheck(Surv2(age, dx2, repeated='first') ~1, ddata, id=ptnum)
check2$transition

# add categorized amlyoid, and initial CMC,
#  but we want lvcf to work correctly wrt amyloid
atemp <- with(ddata, lvcf(ptnum, centiloid))
temp <- ifelse(is.na(atemp), 4,
               as.numeric(cut(ddata$centiloid, c(-50, 25, 68, 500))))
ddata$agrp <- factor(temp, 1:4, c("low", "mod", "high", "miss"))
findex <- match(ddata$ptnum, ddata$ptnum)
ddata$icmc <- ddata$cmc[findex]
boxplot(centiloid ~ agrp, ddata)
abline(h=c(25, 68), col=2)
```

Now fit two models, using CMC and initial CMC.
```{r, mfit1}
mfit1 <- coxph(list(Surv2(age, dx2) ~ male + any.e4 + educ + cmc,
                     1:2 ~ male*any.e4), 
                data= ddata, id=ptnum) 
mfit2 <- coxph(list(Surv2(age, dx2) ~ male + any.e4 + educ + icmc,
                     1:2 ~ male*any.e4), 
                data= ddata, id=ptnum) 
matplot(cbind(coef(mfit1), coef(mfit2)), 1:17)
abline(v=0, lty=3)
```

How do I get a probability in state curve for a 65 year old female, e4-,
cmc of 2, current state of ND, and *current* centiloid of 30?
The standard survfit call doesn't work for TD covariates.
Answer, make more states.

```{r, manystate}
with(ddata, table(agrp, dx2))
temp1 <- with(ddata, lvcf(ptnum, ifelse(dx2=="censor", NA, dx2)))
temp1 <- factor(temp1, 2:4, c("ND", "dementia", "death"))
with(ddata, table2(agrp, temp1))

temp2 <- with(ddata, as.numeric(agrp) + 4* as.numeric(temp1) -4)
dx3 <- factor(temp2, c(0:8, 12), 
              c("censor",
                outer(c("low", "mod", "high", "miss"), c("ND", "dementia"),
                      paste, sep='_'), "death"))
check3 <- survcheck(Surv2(age, dx3) ~ 1, id=ptnum, ddata)
check3$trans

dim(check3$trans)
table(check3$trans[, -10] >0)
table(check3$trans[, -10])
```

This is as far as I got.



