---
title: Practical multistate models
author: Terry Therneau
date: May 30, 2026
output: pdf_document
---

This was prepared as a class session for the advanced survival class in 
Waterloo.

# Introduction

Multistate time to event models based on hazards, i.e., multistate hazard models
(MSH) have been an extremely useful tool for the medical research teams that I
have been involved in at Mayo Clinic. See for instance Jack et al, PMID:35310829
and PMID: 41240917, Allen et al PMID: 41240917 and PMID: 28941364 and 
Le-Rademacher PMID: 36003089 as examples.
They are able to capture multiple aspects of a disease process, and thus can be
an important tool for deeper understanding.

This document is a "nuts and bolts" description that reprises some of those
analyses, with a focus on the data preparation, data checking, and analysis
code needed. Some mathematics will appear, but only to help clarify the
models.

For all the models, key first steps are

 * What data do we have, and what questions *could* be answered with that data?
 * The overall target: what aspects of the process(es) do we wish to
 describe, and in relation to what predictors? No paper can do everything.
 * Lay out the multi-state model as a diagram. Refer to it often.
 * Create the analysis data set, with particular attention to questions of
   immortal time bias.
 * Look at the data (look, not just glance) and decide what to do with 
   special cases. There are invariably one or two.
   
Mathematically we describe out multistate model in terms of three processes.

 * $N_{ijk}(t)$ = the number of observed transitions from state $j$ to state $k$,
   up to time $t$ for subject $i$. 
 * $Y_{ij}(t)$ equals 1 if subject $i$ is currently in state $j$, at time $t$, 
   and at risk for transition to another state at that time. Subjects who have
   entered an absorbing state prior to $t$, e.g., death, are for instance not
   at risk for other transitions. 
 * $X_i(t)$ = the set of covariates for subject $i$ at time $t$. Most often, most
   of these covariates will remain static, e.g., treatment arm or sex, while 
   others are time-dependent, e.g. cholesterol might be remeasured at each visit.

Any careful investigation of the theory behind these models will need to
distinguish between "at risk for an event" and "at risk for an observed event".
In this document I will largely ignore this, and use $Y, N, X$
to refer to observed data.
 If censoring is not random, however, then the two notions are not the same,
estimators can be biased, and any investigation into the hows and whys of that
bias will need to carefully distinguish them.

## Immortal Time Bias

In our opinion, one of the greatest threats to valid medical research is the 
presence of immortal time bias (ITB). In short, for a valid time-to-event
model none of $N(t)$, $Y(t)$ nor $X(t)$ can contain values that depend on the
future. MSH models, unfortunately, provide even more opportunities for
this error to creep in, ones which are sometimes difficult to spot.
The number of ways to create ITB appears to be near infinite. Here are a few
examples.

fill in later

## Timeline data
The survival package supports two forms of multistate data. The older and
better known one will be referred to as counting process form. 
The first 3 lines below show the myeloid data in its raw form as found in
the survival package, followed by the counting process form.
The the tmerge call to create mdata1 will be discussed later.

```{r, mdata1}
library(survival)
library(mshmm)
table2 <- function(...) table(..., useNA="ifany")

# we will show this later
mdata1 <- tmerge(myeloid[,1:4], myeloid, id=id, death= event(futime, death),
                 CR = event(crtime), SCT= event(txtime), relapse=event(rltime),
                 priorcr = tdc(crtime), priorsct= tdc(txtime))

subset(myeloid, id<4)

subset(mdata1, id<4)
```
Each row of data in the timeline form consists of a subject id,
a time interval (tstart, tstop], the event(s) that occur at the end of the
interval (death, complete response, stem cell transplant, relapse), 
and covariates (trt, sex, flt3, priorcr, priorsct).
Covariate apply over the entire interval (tstart, tstop].
Notice the difference between the event variable `CR` and the
time-dependent covariate `priorcr`.
Subject 1 experienced CR on day 44; the CR variable is 1 at the end of that
first (0,44] interval and 0 thereafter. 
The time-dependent covariates priorcr becomes 1 at interval (44,113] 
and remains 1 thereafter. 
The fixed covariates of treatment, sex, and FLT3 repeat from row
to row.

Now the timeline form of these 3 subjects.

```{r, echo=FALSE}
t1 <- with(myeloid, data.frame(id, trt, sex, flt3, time=0))
t2 <- with(subset(myeloid, !is.na(txtime)), data.frame(id, time=txtime,
                                                       sct=1))
mdata2 <- merge(t1, t2, all=TRUE)
t3 <- with(subset(myeloid, !is.na(crtime)), data.frame(id, time=crtime,
                                                       cr=1))
mdata2 <- merge(mdata2, t3, all=TRUE)
t4 <- with(subset(myeloid, !is.na(rltime)), data.frame(id, time=rltime,
                                                       relapse=1))
mdata2 <- merge(mdata2, t4, all=TRUE)
t5 <- with(myeloid, data.frame(id, time=futime, death))
mdata2 <- merge(mdata2, t5, all=TRUE)

indx <- match(mdata2$id, myeloid$id)
for (i in c("trt", "sex", "flt3")) mdata2[,i] <- myeloid[indx, i]
```

```{r, echo=TRUE}
subset(mdata2, id<4)
```

In timeline data each row has a unique (subject identifier, time) pair, along
with all the information that was observed at that time, with missing values
for variables or outcomes that were not observed. 
Timeline data sets have many NA values.
Static variables like treatment, sex and FLT3 state at randomization would 
normally be filled in for every row, as was done here. 
One could argue, however, that FLT3 was
only observed at the start, since a pathology assessment was not repeated,
so should be NA on row 2-4.
Normally, most covariates will be non-missing at baseline, e.g., we know that
CR, SCT, relapse and death have not yet occured so would be 0.
The NA in this case is a function of how we created the data, and will be
handled properly by both Surv2 and lvcf since they are 0/1 (or FALSE/TRUE)
variables.

For simple survival timeline style makes little sense: each subject will have
to have two rows, one at 0 with the covariates and another at the death/censoring
time with all covariates missing.
For more multistate data there may be advantages.

 * Counting process data sets often require the use of tmerge, which
   can be tricky. Timeline data can be built using more standard tools, which
   may be easier.
 * There is only one CR and one SCT variable, not two of each.
   The choice between using a covariates as
   an endpoint or a time-dependent covariate is derferred until is actually
   appears in a model statement, on either the left hand side (as an event) or
   right hand side (as a covariate).
 * If one is rigorous with respect to the "observed at this time" rule for each
   line of data, many immortal time bias problems will be avoided.

# Example 1, NAFLD
The NAFLD data is found in 3 data sets. 
The original study contains all new NAFLD diagnosis in Olmsted County from 
1997 to 2014, along with 4 potential controls for each case. Controls were
selected matched on age, sex, and calendar year.

To protect patient confidentiality all time intervals are in days since the 
index date; none of the dates from the original data were retained. 
Subject age is their integer age at the index date, and the subject 
identifier is an arbitrary integer. As a final protection, the data includes
only a 90% random sample of the original data.

 * nalfd1 has one row for each subject and contains the time fixed variables
 * nafld2 contains time-dependent laboratory tests
 * nafld3 contains the outcomes
 
```{r, nafld1}
nafld1[1:5,]

nafld2[1:5,]

nafld3[1:5,]

table(nafld2$test)

table(nafld3$event)
```

The analysis in Allen 2018 looked at the accumulation of 3 metabolic
comorbidities, leading to the 4 state model shown below.
Age, male sex, NAFLD diagosis, and BMI are the available baseline covariates.
However, BMI is also a major risk factor for development of fatty liver. 
(We didn't include ongoing height/weight in the R data sets.)

```{r, nafld2}
nstate <- c("0MC", "1MC", "2MC", "3MC", "death")
nmat <- matrix(0, 5, 5, dimnames=list(nstate, nstate))
nmat[1,2] <- nmat[2,3] <- nmat[3,4] <- 1
nmat[1:4,5] <- 1
statefig(rbind(4,1), nmat)
```

In the nafld1 data set, the case.id variable, for a control, gives the
id of the case they were matched with;
the incident cases have case.id = id.
The nafld3 data also has nafld diagnosis, so check for consistency.

```{r, nafld3}
ndata <- nafld1  # working data set
ndata$nafld <- 1*(ndata$id == ndata$case.id) # create a variable

table(ndata$nafld)
ndx  <- subset(nafld3, event=="nafld")
table(ndx$days==0)
```

What is going on?

 1. The 318 NAFLD diagnoses after day 0 are subjects who where chosen as 
 controls, who were found to have NAFLD later
 in life. You cannot exclude someone as a control at age $a$ based on something
 that will happen to them some years later at age $a +s$: 
 this would be a form of immortal time bias.
 When that control later became a case, 4 new controls for them were
 selected. 
 2. At this time NAFLD had an interesting definition: 
 if someone had fatty liver but also say a hepatitis C diagnosis,
 then they "didn't have" NAFLD. The study question was thus
 more or less the consquences if NAFLD was the *only* liver disease.
 3. What about hepatitis C 2 weeks later, 6 weeks later, ... The diagnostic
 process for someone with a GI complaint can sometimes stretch out for months,
 so the decision was that "real" NAFLD starts once 12 months has passed 
 without another smoking gun. This is day 0.
 4. Controls need to matched on this new date, otherwise the NAFLD subjects
 will benefit from immortal time bias.
 5. The 19 who have NAFLD at day 0, but are not an incident case, are confusing.
 A of the 19 (not shown) reveals that all were originally chosen as a control.
 Perhaps I made a mistake when going from dates
 to days? (It should be "days from inital recruitment" for all).
 
"If you don't know where you are going, you might end up someplace else". 
Yogi Berra

What should we do? Different choices lead to a different data set and analysis
changes.
 * Simple approach: only follow the orginal enrollees and their controls. Ignore
 when someone changes the time-dependent covariate to NAFLD.
 * Treat NAFLD as a time-dependent covariate
 * Treat NAFLD as a time-dependent state
 * Restart: when a case becomes a control, split them into two copies:
   an original control and a new case.


## Simple approach
For the simple analysis, we need a working data frame that has
a row/days pair each time that the total number of MC increases.
We can take advantage of the fact that each of the MC only is recorded
at most once per subject: formally the variable is "history of diabetes", etc.
The lines below are likely an inefficient way to do this, but
it works.

```{r, nafldssimple}
# first, subset to the relevant cases and controls
caseid <-  subset(ndata, id==case.id)$id
tdata1 <- subset(ndata, case.id %in% caseid)
dim(tdata1) # this is the set subjects, one row each

# the endpoints of interest
enddata <- subset(nafld3, (id %in% tdata1$id) & 
                        (event %in% c("diabetes","dyslipidemia",  "htn")))
enddata[1:10,]
table(enddata$days <=0)
temp <- with(enddata, table(days <=0, event[,drop=TRUE]))
temp
round(temp/nrow(tdata1), 2)

# counting process data
ndata2 <- tmerge(tdata1[c(1,2,3,6:7,10)], tdata1, id=id, 
                 death= event(futime, status),
                 options=list(tstartname="day1", tstopname="day2"))
# Add the MC one by one
ndata2 <- tmerge(ndata2, subset(enddata, event=="diabetes"), id=id, 
                 diabetes= event(days), prior.diab= tdc(days))
ndata2 <- tmerge(ndata2, subset(enddata, event=="dyslipidemia"), id=id, 
                 dyslipid = event(days), prior.dysl= tdc(days))
ndata2 <- tmerge(ndata2, subset(enddata, event=="htn"), id=id, 
                 htn = event(days), prior.htn= tdc(days))

# alternate: add them all at once
ndata2b <- tmerge(ndata2, enddata, id=id, mcount=cumtdc(days), 
                 mevent= cumevent(days))
#
subset(ndata2, id <8)
summary(ndata2)  # a step by step summary of the additions

# Create the combined endpoint
nstate <- c("0MC", "1MC", "2MC", "3MC", "death")
temp1 <- with(ndata2, prior.diab + prior.dysl + prior.htn)
temp2 <- with(ndata2, diabetes + dyslipid + htn)
ndata2$istate <- factor(temp1, 0:3, nstate[1:4])
# temp3 = death(4), or on the day of a new MC, the new cumulative number
temp3 <- ifelse(ndata2$death==1, 4, ifelse(temp2==0, 0, temp2+temp1))
ndata2$event <- factor(temp3, 0:4, c("censor", nstate[-1]))
check1 <-survcheck(Surv(day1, day2, event) ~1, data=ndata2, 
                   id=id, istate=istate)
check1

# This won't be used further, but create the 9 state outcome of Diabetes,
#  dyslipidemia=L and hypertension=H
temp1 <- with(ndata2, prior.diab + 2*prior.htn + 4*prior.dysl)
temp2 <- with(ndata2, diabetes + 2*htn + 4*dyslipid)
e1 <- with(ndata2, paste0(ifelse(prior.diab==1, "D", ""),
                          ifelse(prior.htn ==1, "H", ""),
                          ifelse(prior.dysl==1, "L", "")))
e1 <- ifelse(e1=="", "none", e1)
elab <- c("none", "D", "H", "L", "DH", "DL", "HL", "DHL") #preferred order
ndata2$e1 <- factor(e1, elab)

e2 <- ifelse(ndata2$death==1, 9, ifelse(temp2==0, 0, temp1+temp2))
ndata2$e2 <- factor(e2, c(0, 1,2,4,3,5,6,7,9), c("censor", elab[-1], "death"))
check2 <- survcheck(Surv(day1, day2, e2) ~ 1, ndata2, id=id, istate=e1)
```

For an MSH model fit with survfit or coxph the underlying likelhood is
event based. 
Each observed event contributes a term to the likelihood, and presumes

 * over the window of observation for a subject, all the transitions for
   that subject are recorded
 * the window of observation is complete, i.e., it includes all times at
   which an event, if it occured, would have been recorded
 * non-informative censoring

One consequence of the first is that we do not need to inform the routine
of which transitions can occur, and which are (effectively) absorbing: it
tabulates them on its own.
But, we do need to check that it's assessment agrees with our own. In this
case we have some jumps that were not in our original concept.

```{r, nfig2}
nmat <- matrix(0, 5, 5, dimnames=list(nstate, nstate))
nmat[1,2] <- nmat[2,3] <- nmat[3,4] <- 1
nmat[1:4,5] <- 1
nmat[1, 3:4] <- nmat[2,4] <- .7
statefig(rbind(4,1), nmat, alty=c(1,2,1,2,2,1,1,1,1,1))
```

To properly account for age we will do the primary analysis on age scale.

```{r, nafldage}
quantile(tdata1$age)
ndata2$age1 <- ndata2$age + ndata2$day1/365.25
ndata2$age2 <- ndata2$age + ndata2$day2/365.25

naj <- survfit(Surv(age1, age2, event) ~ male, ndata2, id=id, istate=istate,
               p0= c(1,0,0,0,0), start.time=50)

dim(naj)
plot(naj[1,2:5], lwd=2, col=1:4, xmax=100,
     xlab="Age", ylab="Probability in state")
plot(naj[1,], cumprob=2:5, col=1:4, yaxs='i', xmax=100,
     xlab="Age", ylab="Probability in state")
#plot(naj, cumprob=2:5, lty=1:2,  col=rep(1:4, each=2), yaxs='i', xmax=100,
#     xlab="Age", ylab="Probability in state")
text(c(55, 68, 77, 79, 89), c(.8, .11, .37, .57, .69), nstate,
     col=c(1,1,2,3,4))

dim(naj$cumhaz)
plot(naj[1,], cumhaz=TRUE, col=1:5, lty=1:2, 
     xlab="Age", ylab="Cumulative Hazard")
```
The first two plots show probability in state, for females only,
in two different forms
Some users prefer the first, some the second.
The second figure is closely related to a stacked bar chart; the probability in
state is the difference between two curves.
The third plot shows the 10 separate estimated hazard curves.
The probability in state estimates are mostly oblivious to the presence 
of  0mc:2MC, 0MC:3MC and 1MC:3MC transtions, but the hazard is estimated
separately for each pair, some of which have small numbers.
This affects the multi-state Cox model as well, since it is focused on hazards.

```{r, nacox1}
nacox1 <- coxph(Surv(age1, age2, event) ~ nafld + male,
                data=ndata2, id=id, istate=istate)
nacox1$cmap
```

The simple Cox model failed to converge. The coefficient map showed that 
coefficient 8 is associated with the 1:4 transition, which involves
only a handful of subjects.
To ameliorate this, bundle the coeffients for the small transitions.

```{r}
nacox2 <- coxph(list(Surv(age1, age2, event) ~ nafld + male,
                     1:2 + 1:3 + 1:4 ~ nafld + male /common,
                     2:3 + 2:4 ~ nafld + male /common,
                     0:5 ~ male/common),
                data=ndata2, id=id, istate=istate)
nacox2

round(coef(nacox2, matrix=TRUE),3)
```

Because all the covariates are time fixed, absolute risk estimates are
fairly easy to get. For a Cox model, predictions are available for any
a priori combination of covariates, which we provide in a dummy data set.
Since it is multistate, we need to give a distribution of starting
states, and since we are on age scale a starting time.
(If you don't give these it will guess. I don't think I have every found the
result of that guess to be useful.)

```{r, absnafld}
dummy <- expand.grid(male=0:1, nafld=0:1)
coxsurv2 <- survfit(nacox2, newdata=dummy, p0=c(1,0,0,0,0), start.time=45)
dim(coxsurv2)
print(coxsurv2[1,], rmean=100)

xx <- paste0("male=", dummy$male, ", Nafld=", dummy$nafld)
sojourn <- summary(coxsurv2, rmean=100)$table
sojourn <- matrix(sojourn[,"rmean"], ncol=4, byrow=TRUE,
               dimnames=list(nstate, xx))
rbind(sojourn, Total= colSums(sojourn))


plot(coxsurv2[,-1], col=1:4, lty=rep(1:4, each=4))
```

* The mean time in state has to add to (100-45)= 55, across the states.
* Subscripting the curves is a useful tool.
* The simple plot has too many curves, it looks like a bowl of spaghetti.
  Better plotting requires artistic choices.
* For the stratified model, the defalt plot will have a set of curves per
  stratum, `r length(unique(ndata$case.id))` times 5 states, which is 
  obviously not useful.

```{r, absnafld2}
opar <- par(mfcol=c(2,2), mar=c(5,5,2,1))
xx <- outer(c("F", "M"), c("No NAFLD", "NAFLD"), paste)
for (i in 1:4) {
    plot(coxsurv2[i,], cumprob=2:5, col=1:4, lty=rep(1:2, each=4))
    text(55, .9, xx[i])
}
par(opar)

barplot(sojourn[,c(2,4,3,1)], horiz=TRUE, col=c(5:2, 0), las=2,
        names.arg=c("Male", "MaleN", "FemaleN", "Female"),
        xlab="Years in state", border=NA, legend=TRUE)
```


# Mayo Clinic Study of Aging

The MCSA is based on a stratified random sample of subjects from Olmsted
County, Minnesota. 
The dementia data set in the mshmm package is based on this data, and is
able to reproduce Jack et al (2022) quite well, so we will use it for analysis.
This data set is already in timeline form.

```{r, mcsa1}
mstate <- c("CU", "MCI", "dementia", "death")
mmat <- matrix(0, 4, 4, dimnames=list(mstate, mstate))
mmat[1,2] <- mmat[2,3] <- 1
mmat[1:3,4] <- 1
statefig(rbind(3,1), mmat)
```

```{r, mcsa2}
survcheck(Surv2(age, dx, repeated=TRUE) ~1, id=ptnum, dementia)$transitions

survcheck(Surv2(age, dx) ~1, id=ptnum, data = dementia)
```
We can see several things by looking at this table.

  * The survcheck routine was written for counting process data.
    One aspect of Surv2, which translates from timeline form to that form,
    is the repeated argument.  When a subject has two sequential visits in
    the same state, the survfit and coxph routines need to know whether to
    treat the second observation as a new event. Most of the time they
    should not: if "diabetes" is recorded in the medical history for 5 visits
	in a row this is not 5 new instances of the condition.
	Repeated infections in the CGD data set is a counterexample, however.
	In this data each row is a visit, the repeated form clearly shows
    that over half of all visit pairs are CU:CU; most patients stay stable for
    many years before progression.  The (any) row of the total visits table
    shows that 53\% (2296/4327) have no recorded transtions during follow-up.
  * There are 30 instances of a CU:dementia pair.  Per the model we assume
    that these subjects passed through MCI, we simply did assess them often
    enough to observe the transient MCI state.
  * Dementia:MCI is much less common than MCI:CU.  This reflects on one
    hand the clinical reality that CU and MCI are more difficult to distinguish
    than dementia, and also fewer overall dementia cases.
  * There are no transitions from death to another state, or from
    death to censored (further follow-up after death).  This is as it
    should be.
  * Off study dementia only transitions to death. This is because
    further record reviews are not scheduled for demented subjects.
  * There are no transitions from nondem to nondem or off-dem. 
    This is, it turns out, an interesting artifact of how the online
    data collection system was designed.
    The nurse abstracters who review the history have a screen with fields
    for ``last assessment date for dementia' and 'dementia yes/no'.
    At a second, third, etc review for a patient, should one occur,
    the contents of these are overwritten; thus the master data set
    at any given time only has the most recent value.
  * More dementias were found in record review than in study visits.
    We believe that this reflects informative censoring: subjects whose
    cognition is slipping selectively drop out, i.e., exactly those most
    likely to receive a dementia diagnosis at their next visit. 

Jack et al (2022, PMID: 35310829) looked at a model for prediction of the
absolute risk of dementia, 
a key question was the additional predictive power of amyloid PET after 
accounting for known risk factors of age, sex, the presence of one or more
e4 allelles on the APOE gene, education, and CMC score (a count of
7 cardio-metabolic conditions).
It used the modified endpoint of non-demented/dementia/death, collapsing the
CU and MCI categories.
First create the modified diagnosis variable:

```{r, nondem}
dem <- dementia
table(dem$dx)
temp <- c(1, 2,2,3,2,3,4)[as.numeric(dem$dx)]
dem$dx2 <- factor(temp, 1:4, c("censor", "ND", "dementia", "death"))
with(dem, table2(dx2, dx))

survcheck(Surv2(age, dx2) ~1, dem, id=ptnum)
```

The above shows 29 transtions from dementia to ND.
The 2022 analysis was actually for time to *first* dementia. This change to
using only first event is common enough that we have added a routine to
the survival package.
```{r, first}
# the next line doesn't yet work though
# dx3 <- with(dem, nostutter(ptnum, dx, censor="alive", single=TRUE))
# do it with unlist(tapply(...  which is very obscure
tfun <- function(x) { y <- as.numeric(x); ifelse(y==1, 1, cummax(y))}
temp <- with(dem, unlist(tapply(dx2, ptnum, tfun)))

dem$dx3 <- factor(temp, 1:4, c("censor", "ND", "dementia", "death")) 
dem$edu4 <-pmax(8, dem$educ) /4
check2 <- survcheck(Surv2(age, dx3) ~1, dem, id=ptnum)
check2$transition
```

Amyoid PET raises some important statistical issues.
First is that only about 40% of the subjects have amyloid imaging.
This appears to be largely a time issue: imaging is a second appointment of
2 full days (MRI, amyloid PET, tau PET, optional FDG PET), and many who have
previously given informed consent for imaging defer at this point.
A second is that the ligand used for amyloid PET did not become available
until several years after the start of the MCSA, so many of those who have
imaging do not start at visit 1. 
Third is that the pattern of amyloid accumulation has an elbow, and each subject
has a unique age for their bend.
Some prior work had categorized amyloid as low/moderate/high; our choice in
this study was to use a time-dependent category of low/moderate/high/missing.

```{r, data}
# add categorized amlyoid, and initial CMC,
#  but we want lvcf to work correctly wrt amyloid, apply it before making NA
#  a category
atemp <- with(dem, lvcf(ptnum, centiloid))
temp <- ifelse(is.na(atemp), 4,
               as.numeric(cut(atemp, c(-50, 25, 68, 500))))
dem$agrp <- factor(temp, 1:4, c("low", "mod", "high", "miss"))
firstrow <- match(dem$ptnum, dem$ptnum)
dem$icmc <- dem$cmc[firstrow]
boxplot(centiloid ~ agrp, dem)
abline(h=c(25, 68), col=2)
```

Now fit two models, using CMC and initial CMC.
```{r, mfit1}
mfit1 <- coxph(list(Surv2(age, dx3) ~ male + any.e4 + cmc + agrp,
                     1:2 ~ male*(any.e4 + agrp)), 
                data= dem, id=ptnum) 
mfit2 <- coxph(list(Surv2(age, dx3) ~ male + any.e4 + icmc + agrp,
                     1:2 ~ male*(any.e4 + agrp)), 
                data= dem, id=ptnum) 

opar <- par(mar=c(5,6,1,1))
matplot(cbind(coef(mfit1), coef(mfit2)), 1:22, xlab="Coefficient",
        ylab="", yaxt='n')
abline(v=0, lty=3)
axis(2, 1:22, names(coef(mfit2)), las=2)
par(opar)
```

Not done: create my own version of Emily's nice plot (below) and insert it
instead.

```{r, echo=FALSE}
knitr::include_graphics("figures/fig2.pdf")
```

How do I get a probability in state curve for a 65 year old female, e4-,
initial CMC of 2, current state of ND, and *current* centiloid of 30,
when centiloid is a time dependent covariate?
The standard survfit call (curves for a coxph model)
doesn't work for TD covariates.
For CMC, we are able to dodge the issue by instead using a time-fixed variable of
initial CMC at recruitment.
The plot above shows that hazard ratios don't change very much.
But that is not an option for amyloid: any "initial amyloid" variable that we
try to create leads to immortal time bias.


One approach is to make more states,
creating 4 amyloid * (ND, dementia) + death= 9 states.
One could also have 4*3 = 12, separating out death with low amyloid, death
with moderate amyloid, etc.
(The reason to use initial CMC instead of time-dependent CMC is so that this
example has a more managable number of states.  The plot above shows that the
models with initial and time dependent CMC are quite similar.)

Create this new set of states. 
Code wise, it was easiest to first fill out the censored
values in the `dx3` variable to make it complete.
We also need to create the first dementia status ourselves, as the automated
methods may not recognize a mod/dementia to mod/ND transition
as a should-not-be-allowed repeat of ND.

```{r, crossed, echo=FALSE}
# for the graph and text, it works better to put missing on the left
sname <- c(outer(c("miss", "low", "mod", "high"), c("ND", "dem"), paste, 
                 sep='/'), "death")
alias <- data.frame(state= sname, 
                    A= c(1:4, 1:4, 0),
                    C= c(rep(1:2, each=4),3))
# for the graph, put missing amyloid on the left (simpler writing)
Qmat <- qplus(alias, death="death")

# for statefig, reformat the state names to two lines
sname2 <- c(outer(c("miss", "low", "mod", "high"), c("ND", "dem"), paste, 
                 sep='\n'), "death")
tempq <- Qmat
dimnames(tempq) <- list(sname2, sname2)
statefig(rbind(4,4,1), Qmat, acol=rep(c("black", "gray"), c(10,8)))
```


```{r, manystate}
with(dem, table(agrp, dx3))
temp1 <- with(dem, ifelse(dx3=="censor", NA, as.numeric(dx3)))
temp2 <- with(dem, lvcf(ptnum, temp1))
temp2 <- factor(temp2, 2:4, c("ND", "dementia", "death"))
with(dem, table2(agrp, temp2))

temp3 <- with(dem, as.numeric(agrp) + 4* as.numeric(temp2) -4)
# the pmin collapses 9-12= low/death, ..., miss/death in to "death"
dem$dx4 <- factor(pmin(9, temp3), 0:9, 
              c("censor",
                outer(c("low", "mod", "high", "miss"), c("ND", "dem"),
                      paste, sep='/'), "death"))
check3 <- survcheck(Surv2(age, dx4) ~ 1, id=ptnum, dem)
check3$trans
```

```{r, echo=FALSE}
# add all the other arrows
tempx <- tempq
tempx[1, 2:4] <- .6
tempx[3,2] <- tempx[4,3] <- 1
tempx[2,4] <- .6
tempx[1,6:8] <- 1
tempx[2,7] <- 1
tempx[6,8] <- .6
statefig(rbind(4,4,1), tempx, acol=rep(c("black", "gray"), c(21,8)))
```

Writing a coxph call is going to be a bother: there are 9 transitons that 
have a count of 5 or less, all of which would need to be bundled with some
other one(s) to avoid undefined coefficients.
Take another approach. Start with the simpler figure.

For the MSH model, I am interested in only the downward arrows.
One could explore the other factors, if any, the affect amyloid progression,
i.e., rightward arrows from low to moderate or moderate to high,
but the focus of the 2022 paper was on the prediction of cognitive state.
Transitions from missing amlyoid to low/moderate/high amyloid, 
lines 4 and 8 of the transitions table, occur for subjects
who entered the study before the PET ligand for amyloid 
(Pittsburg compound B) became available, those transitions are based
on calendar date of the visit and have nothing to do with disease per se.

An alias data set is useful here; it allows shorthand formulas
within the coxph and cmsh functions.

```{r, multifit1}
alias

mfit3 <- coxph(list(Surv2(age, dx4) ~ 1 ,
                   C(1):C(2)    ~ male*(any.e4 +agrp)+ icmc + 1/ common,
                   C(1):"death" ~ male + any.e4 + icmc + agrp + 1 /common,
                   C(2):"death" ~ male + any.e4 + icmc + agrp + 1 /common),
              data=dem, id=ptnum, statedata= alias)
mfit3
```

In a Cox model, either single endpoint or a MSH, the baseline hazard plays
a similar role to the intercept in a GLM, but it's estimation can be
deferred until after solving for the coefficients $\beta$; it is similar
in that way to the estimation of $\hat\sigma$ in a least-squared model.
In the code above the combination of `agrp + 1` and `\common` on line
1 of the formula signals that proportional baseline hazards will be used
for the set of ND:dementia transitions, i.e. assume hazards of
$$ \begin{aligned}
	&\lambda_{ND, dem}(t) \\
	&\lambda_{ND, dem}(t) e^{\gamma_1}\\
	&\lambda_{ND, dem}(t) e^{\gamma_2}\\
	&\lambda_{ND, dem}(t) e^{\gamma_3}
\end{aligned}
$$
for transitions from the low/ND to any of the 4 dementia states,
for transitions from mod/ND  ($\gamma_1$) to any of the 4 demenetia states,
for transitions from high/ND ($\gamma_2$) and for transtions from 
miss/ND ($\gamma_3$), respectively.
(The $\gamma$s are in the order of the factor levels for agrp, not the
order of my picture).
This includes for instance the one diagnonal transition from 
low/ND to mod/dem found in the transitons table. 
These three $\gamma$ coefficients appear as `agrp` in the first block 
of printed coefficients for the model.

In mfit3, amyloid is no longer a time dependent variable but rather a 
state. There in fact no time-dependent coeffcients, and so we are able
to create honest predictions based on baseline covariates, starting
age, and starting state.

```{r, multifit2, eval=FALSE}
dummy1 <- expand.grid(male=0:1, any.e4 = 0:1, icmc=1)

#msurv3.1 <- survfit(mfit3, newdata = dummy1, start.time=65,
#                    p0=c(1,0,0,0,0,0,0,0,0))
# the above call gives an error
# this is a flaw in the survival package, which is being addressed
```

What is most interesting is that the log-likelihood and coefficients from
`mfit2` and `mfit3` are identical. This is because the underlying partial
likelihood terms, one for each transition with covariates, have exactly
the same set of subjects at risk, with the same covariate pattern.
The coefficient's names and order differ, but values agree.

```{r, same}
rbind(mfit2= mfit2$loglik, mfit3= mfit3$loglik)
```

One of the challenging aspects of multistate models is managing the total
number of parameters. There are 35 separate transitions in this example,
the number of non-zero elements in the transitons matrix. Without constraints,
each transition will be assigned its own coefficient values for male, any.e4,
cmc and amyloid, $6*35= 210$ coefficients, along with 35 separate baseline
hazard functions.
One way to think of this is that the coxph function creates transition *
covariate interactions by default; the same combinatorial explosiion would
occur in linear models if we included all possible interactions.
